---
title: "BI-PRS Semestral Project 2"
authors:
  - Oleksandr Slyvka
  - Illia Lyhin
  - Maksym Khavil
output: html_document
date: "2025-04-20"
---

# Semestral Project 2

Our representative is Oleksandr Slyvka, his M-value is 6, that corresponds to scores in SAT tests, thus we will investigate its data.

## Libraries & Data

```{r setup, include=FALSE}
library(dplyr)
library(vtable)
library(ggplot2)
library(cowplot)
library(corrplot)
library(e1071)
library(tidyr)
library(plotrix)

library(Sleuth2)
```

Load the dataset.

```{r}
data("case1201")
print(case1201)
```

## Data Description & Graphs

A comprehensive excerpt from the [manual](https://cran.r-project.org/web/packages/Sleuth2/Sleuth2.pdf):\
A data frame with 50 observations on the following 8 variables.

1.  **State** US state

2.  **SAT** state averages of the total SAT (verbal + quantitative) scores

3.  **Takers** the percentage of the total eligible students (high school seniors) in the state who took the exam

4.  **Income** the median income of families of test–takers (in hundreds of dollars)

5.  **Years** the average number of years that the test–takers had formal studies in social sciences, natural sciences and humanities

6.  **Public** the percentage of the test–takers who attended public secondary schools

7.  **Expend** the total state expenditure on secondary schools (in hundreds of dollars per student)

8.  **Rank** the median percentile ranking of the test–takers within their secondary school classes

We will need a categorical variable so let's transform **Expend**:

1.  **Low** – expenditure \< \$2000 per student

2.  **Medium** – expenditure [\$2000, \$2500) per student

3.  **High** – expenditure \>= \$2500 per student

```{r}
case1201$ExpendCat <- cut(case1201$Expend,
                          breaks = c(-Inf, 20, 25, Inf),
                          labels = c("Low", "Medium", "High"),
                          right = FALSE)
table(case1201$ExpendCat)

```

Let's display some basic descriptive statistics for each column.

```{r}
case1201_numerical_cols <- subset(case1201, select = -c(State, ExpendCat))
summ <- c('mean(x)', 'sd(x)', 'skewness(x)', 'min(x)', 'pctile (x)[25]', 
          'median(x)', 'pctile(x)[75]', 'max(x)', 'IQR(x)'
)

sumtable(
  case1201_numerical_cols, colnames(case1201_numerical_cols),
  out='kable',
  summ=summ
)
```

Let's explore relationships between variables using correlation matrix:

```{r}
numeric_data <- case1201[sapply(case1201, is.numeric)]

cor_matrix <- cor(numeric_data, use = "complete.obs")

corrplot(cor_matrix, addCoef.col = "goldenrod4", tl.col = "black")
```

Our correlation insights might be:

1.  Higher test participation is linked to lower average SAT scores.

2.  States with higher income and better educational rankings and higher **Years** variable tend to have higher SAT scores.

3.  Variables like Public school percentage and Expenditure per student show weak or negligible correlations with SAT scores.

Let's visualize the distribution of the most relevant variables in the dataset.

```{r}
hist(case1201$SAT,
     main = "Distribution of SAT Scores",
     xlab = "SAT Score",
     col = "lightblue", border = "white")

hist(case1201$Takers,
     main = "Distribution of SAT Takers",
     xlab = "Number of Takers (%)",
     col = "lightgreen", border = "white")

hist(case1201$Income,
     main = "Distribution of Median Income",
     xlab = "Income ($100s)",
     col = "lightcoral", border = "white")

hist(case1201$Rank,
     main = "Distribution of Educational Rank",
     xlab = "Rank",
     col = "plum", border = "white")

```

Let's take a quick look on scatter plots and Pearson correlation coefficient to make some assumptions about linear trends or clusters between SAT scores and other variables.

```{r}
plot(case1201$Takers, case1201$SAT,
     main = "SAT Score X % of Test Takers",
     xlab = "Takers (%)",
     ylab = "SAT Score",
     col = "darkgreen", pch = 19)

plot(case1201$Income, case1201$SAT,
     main = "SAT Score X Median Income",
     xlab = "Income ($100s)",
     ylab = "SAT Score",
     col = "orangered", pch = 19)

plot(case1201$Rank, case1201$SAT,
     main = "SAT Score X Secondary School Rank",
     xlab = "Rank",
     ylab = "SAT Score",
     col = "purple", pch = 19)

plot(case1201$Years, case1201$SAT,
     main = "SAT Score X Average Years of Education",
     xlab = "Years of Education",
     ylab = "SAT Score",
     col = "orange", pch = 19)

```

```{r}
print("Correlation with Sat")
cor_with_sat <- cor_matrix["SAT", ]
print(cor_with_sat)
```

Based on the plots above and strong enough positive correlation coefficient **(0.88)** we chose **Rank** as a regressor for the second task.

## Numerical Regressor

We will compare the linear and quadratic regression models to determine which better explains the relationship between SAT and Rank.

```{r}
model_lin <- lm(SAT ~ Rank, data = case1201)
summary(model_lin)
```

```{r}
model_quad <- lm(SAT ~ Rank + I(Rank^2), data = case1201)
summary(model_quad)
```

Let's plot the regression line and curve so we have a broader picture.

```{r}
plot(case1201$Rank, case1201$SAT,
     main = "SAT vs. Rank with Linear and Quadratic Fit",
     xlab = "Rank", ylab = "SAT Score",
     pch = 19, col = "gray")

abline(model_lin, col = "blue", lwd = 3)

x_vals <- seq(min(case1201$Rank), max(case1201$Rank), length.out = 100)
y_vals_quad <- predict(model_quad, newdata = data.frame(Rank = x_vals))
lines(x_vals, y_vals_quad, col = "red", lwd = 3)

legend("bottomright",
       legend = c("Linear Fit", "Quadratic Fit"),
       col = c("blue", "red"), lwd = 2)
```

Now we have to choose which model to use. **Linear Model** is great, mainly because it's interpretation is very clear: *for each 1-point increase in Rank, the SAT score increases by \~9.56 points*. *Intercept means that the SAT score of person with 0 point Rank may be predicted as \~183 points.*

Both of the models explain high portion of variability in data, however **Quadratic Model** has a slightly better fit (Adjusted R² = 0.779 vs. 0.769). None of its coefficient are statistically significant, so despite having better adjusted R² it should be reducible to **Linear Model**. We can use ANOVA which helps us to test:

$$
H_0: \text{The additional predictors in Quadratic Model do not improve the model.}\\
H_A: \text{At least one of the additional predictors in Quadratic Model has a non-zero coefficient.}
$$

```{r}
anova(model_lin, model_quad)
```

Under significance level of $α=0.05$ we fail to reject the null hypothesis. So we can choose **Linear Model**.

## Categorical Regressor

**ExpendCat** will be used as categorical regressor. First we will apply analysis of variance on it to compare the means of SAT score in different categories. So let's examine SAT score distributions in each category:

```{r}
boxplot(SAT ~ ExpendCat, data = case1201,
        main = "SAT Scores by Expenditure Category",
        xlab = "Expenditure Category", ylab = "SAT Score",
        col = c("lightblue", "lightgreen", "lightcoral"))
```

It is hard to tell if the means are different, **High** category only shows, that disribution there is skewed. So we will apply analysis of variance on **ExpendCat** to test:

$$
H_0: \text{The mean SAT scores are the same across all categories.}\\
H_A: \text{At least two ExpendCat categories have different SAT scores means.}
$$

```{r}
anova(aov(SAT ~ ExpendCat, data = case1201))
```

The *p-value* is much greater than $0.05$, thus the result is not statistically significant and we fail to reject the null hypothesis. The ANOVA test shows that **ExpendCat** does not significantly affect SAT scores (SAT scores do not differ meaningfully between low, medium, and high-expenditure groups).

Next we will see how bad will be predictions with such regressor.

```{r}
fit=lm(SAT ~ ExpendCat, data = case1201)
summary(fit)
```

Coefficients:

-   **Intercept**: This is the average SAT score for states (944.94) in the Low expenditure category.

-   **ExpendCatMedium**: On average, SAT scores in the Medium category are only \~11 points higher than in the Low category, but this difference **is not statistically significant**.

-   **ExpendCatHigh**: SAT scores in the High expenditure category are very slightly lower (by \~1 point) than the Low category, **and this is also not significant**.

So the model also suggests that total state expenditure on secondary schools has no real effect on average SAT scores. Also the model itself is poor, **Adjusted R-squared is negative**, meaning the model performs worse than a mean-only model.

## Interaction Model

Now we use both of the above regressors together.

```{r}
interaction_model <- lm(SAT ~ Rank * ExpendCat, data = case1201)
summary(interaction_model)
```

Coefficients now are a bit harder to interpret:

-   **Intercept**: Baseline SAT score (–139.45) when Rank = 0 and ExpendCat = Low. However score can not be negative.

-   **Rank**: For Low expenditure states, each additional point in Rank increases SAT by \~13.1 points.

-   **ExpendCatMedium**: Medium-expenditure states have \~289 points higher SAT than Low-expenditure ones if Rank = 0 (intercept shift).

-   **ExpendCatHigh**: High-expenditure states have \~368 points higher SAT than Low-expenditure ones if Rank = 0.

-   **Rank:ExpendCatMedium**: The positive effect of Rank is slightly weaker (–3 pts) in Medium group compared to Low.

-   **Rank:ExpendCatHigh**: The effect of Rank is even weaker in High group (–3.94 pts compared to Low).

Statistical Significance:

-   **Rank** alone is the most significant one component here.

-   **ExpendCat (Medium and High)** significantly shifts the baseline SAT score up (with Rank = 0), though the intercept itself not much meaningful in isolation.

-   **Interaction with High** is clearly significant, confirming that expenditure level modifies the influence of Rank on SAT.

-   **Interaction with Medium** is marginally significant.

**Adjusted R²** is **0.868**, meaning that this model gives a **better explanation** of SAT variation than linear regression with **Rank** alone.

Let's visualize regression lines for different **ExpendCat** categories.

```{r}
ggplot(case1201, aes(x = Rank, y = SAT, color = ExpendCat)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x,se = FALSE) +
  labs(title = "Interaction: SAT ~ Rank * ExpendCat",
       x = "Educational Rank", y = "SAT Score")

```

We see how much scores differ for **Low** category.

## Model with 3 Regressors

```{r}
full_model <- lm(SAT ~ Rank * ExpendCat *Years, data = case1201)
summary(full_model)
```
