---
title: "BI-PRS Semestral Project 1"
authors:
  - Oleksandr SLyvka
  - Illia Lyhin
  - Maksym Khavil
output: html_document
date: "2025-03-11"
---

# Semestral Project 1

Our representatice is Oleksandr Slyvka, his M-value is 6, that corresponds to Spain, thus we will investigate its data.

## Libraries & Data

The following libraries are used in this R Markdown notebook.
```{r setup, include=FALSE}
library(eurostat)
library(dplyr)
```

We will download dataset, further comments are based on [metdata provided by eurostat]()
```{r}
# Raw data
data <- get_eurostat("nama_10r_3empers")
# Selection 2022 only
data <- data[(as.Date('2022-01-01') <= data$TIME_PERIOD) & ( data$TIME_PERIOD < as.Date('2023-01-01')),]
# Number of employed per a thousand citizens
data <- data[(data$wstatus == 'EMP'),]
# Only NUTS 3 regions
data <- data[nchar(data$geo) == 5,]
# Only in Spain
data <- dplyr::filter(data, grepl('ES', geo))
data
```

## Description and Graphs
TODO

## Contingency table
TODO

## Tests
TODO

## Conclusion
TBA